Argumentation theory for mathematical argument\thanks{Support from the UK Engineering and Physical Sciences Research Council is acknowledged under grants EPSRC EP/K040251/2 (Corneli, Martin and Nesin), EP/J017728/2 (Murray-Rust) and EP/P017320/1 (Pease).}

### 2018-07-11 main.tex
update some of the references up Martin and Pease
last reference checked

\newpage 
\paragraph{Our method for producing this set of tags was as follows.}
Two of us (with first degrees respectively in Mathematics and Information
Systems, both with more than 10 years experience studying
argumentation and social machines) performed close content analysis
\cite{klaus2004content} together on the first 100 comments in MiniPolymath
1.  Our analyses resulted in an initial tag set, including both typical
illocutionary performatives and mathematics specific performatives,
like \texttt{Define} and \texttt{QueryE}, as
needed (see Appendix \ref{app:reference-coding-samples} for examples).  Several
of the typical illocutionary connections (\texttt{Assert}, \texttt{Question}, \texttt{Challenge},
\texttt{Agree}) could be carried over from the schemes commonly applied in
IAT.  Our initial tag set was discussed and iteratively developed over
the same 100 comments by all co-authors, with any recurring
differences discussed, allowing us to align our results.  A third
co-author (with a first degree and PhD in Mathematics) then further
developed and finalised the tag set by performing close content
analysis on the entire MiniPolymath 3 conversation and on sections of
MiniPolymath 1.  Again, this was conducted alongside discussion with
the other co-authors throughout the process.
A fourth co-author  (with a first degree in Mathematics)
later extended the tag set
with
additional informal logical relationships, such as \texttt{analogy}, and specific
content-focused relationships, such as \texttt{sums}, which played a
role in the further examples we treated in Section
\ref{iatc-examples}.  These extensions were again reviewed by all co-authors.

@@

The data that underlie this section were generated in a series of online experiments in collaborative problem solving convened by mathematician Terence Tao.\footnote{\url{https://terrytao.wordpress.com/2011/06/09/mini-polymath-3-2011-imo-question/}}\textsuperscript{,}%
\footnote{\url{https://polymathprojects.org/2011/07/19/minipolymath3-project-2011-imo/}}
\thesis{We use IATC to expand on the analysis of
  this data presented by \citet{pease-and-martin},} showing how IATC can
advance the theory of mathematical argument through the
detailed analysis of real world examples, as per \citet{Carrascal2015}.

The earlier analysis by Pease and Martin proposed a typology of comment types, based on the comments' contents:
\emph{examples},
\emph{conjectures},
\emph{concepts},
\emph{proofs},
and \emph{other}.  
An IATC analysis of the same data gives a more detailed picture of the anatomy of the comments that fall within these intuitive categories.  We use this to build a relatively detailed understanding of what went on in the process of constructing this interesting collaborative proof.

@@

\paragraph{Our main example in this section is MiniPolymath 3 (MPM3),
which we tagged into IATC in its entirety.}
As an indicative sample, the first three comments and their tags are shown in Figure
\ref{fig:mpm3-example}.
Figure \ref{fig:grammar-categories} shows how tags from IATC's five grammatical
categories were distributed over time.  Thus, for example, we see
`\texttt{value}' tags used early in the discussion as strategies are
being considered, and again later in the discussion when solutions are
being vetted.
Figure \ref{fig:alternative} gives another view of the timeline,
showing how the comments were categorised into the 5-part typology
from Pease and Martin.  In the initial
categorisation developed for that paper, comments were allowed to be
in multiple categories at once.
\paragraph{Here, to facilitate a clean mapping to IATC, we redid the categorisation with the requirement that each comment should fit into exactly one main category.}
We arrived at a nearly equal division of comments among the five categories:
\emph{example} (20.3\%) \emph{conjecture} (21.2\%),
\emph{concept} (19.5\%), \emph{proof} (19.5\%), and \emph{other}
(19.5\%).

@@

\paragraph{One might suspect that \texttt{Suggest} should be used only within conjectures, but in the current categorisation it is used somewhat more frequently along with concepts.}
This is partly explained by the fact that \texttt{Suggest} can be used
to introduce either a \texttt{goal} or a \texttt{strategy}.  Sometimes goals
represent conceptual tidying, as in ``I guess there is an odd / even
number of point distinction to do'' \cite[\href{https://wp.me/pAG2F-41\#comment-3398}{July 19, 9:31 pm}]{tao2011imo}.

\paragraph{In spite of our self-imposed constraint to map each comment to exactly one of the five categories, in practice a comment may simultaneously introduce a concept along with a conjecture that applies that concept.}
For example the straightforward concept of
``restriction[s] on how the next pivot is chosen''
appears along with the more speculative conjecture  ``Can we
start with a complete graph and all cycles on that graph and just
discard the ones that don't follow the restrictions to converge on the
ones that do?''
\cite[\href{https://wp.me/pAG2F-41\#comment-3368}{July 19, 8:56 pm}]{tao2011imo}.
The need to introduce concepts also applies in the case of more outlandish conjectures,
such as
``It might be fun to use projective duality''
\cite[\href{https://wp.me/pAG2F-41\#comment-3324}{July 19, 8:23 pm}]{tao2011imo}.
However, a concept may suggest a vague method
without raising a conjecture as such,
e.g., ``I'm thinking spirograph rather than convex hull''
\cite[\href{https://wp.me/pAG2F-41\#comment-3349}{July 19, 8:44 pm}]{tao2011imo}.

@@

\paragraph{There are other resources available which could further expand IATC's offerings in this regard.}
For example, a recent special issue of \emph{Argument \& Computation} \cite{ac-special-issue} includes papers detailing the usefulness of
rhetorical structures for argument mining.   \citet{mitrovic2017ontological}, in that volume, indicate the SALT Rhetorical Ontology  \cite{handschuh2006salt,groza2012advances} as relevant prior work.  SALT contains three categories---coherence relations, argument scheme relations, and rhetorical blocks---each of which unfolds with considerable further detail.  These three categories can be seen as somewhat analogous to IATC's grammatical categories.   \citet{mitrovic2017ontological} and \citet{lawrence2017harnessing} point to foundational work of \citet{fahnestock1999rhetorical,fahnestock2004figures} on the argumentative function of rhetorical figures, particularly in science writing.  IATC might be profitably connected to such analyses. Furthermore, the integration of rhetoric into argument mining highlights the relevance of structures that are rather different from the IAT-style transitions that have been used in work summarised by \citet{budzynska2015automatically}.  White's \citeyearpar[p.~6]{white1978tropics} pithy assertion that ``logic itself is merely a formalization of tropical strategies'' can serve as an additional provocation to develop structural analyses of this sort.

>>>>

\newpage 
\paragraph{Our method for producing this set of tags was as follows.}
Two of us (with first degrees respectively in Mathematics and Information
Systems, both with more than 10 years experience studying
argumentation and social machines) performed close content analysis
\cite{klaus2004content} together on the first 100 comments in MiniPolymath
1.  Our analyses resulted in an initial tag set, including both typical
illocutionary performatives and mathematics specific performatives,
like \texttt{Define} and \texttt{QueryE}, as
needed (see Appendix \ref{app:reference-coding-samples} for examples).  Several
of the typical illocutionary connections (\texttt{Assert}, \texttt{Question}, \texttt{Challenge},
\texttt{Agree}) could be carried over from the schemes commonly applied in
IAT.  Our initial tag set was discussed and iteratively developed over
the same 100 comments by all co-authors, with any recurring
differences discussed, allowing us to align our results.  A third
co-author (with a first degree and PhD in Mathematics) then further
developed and refined the tag set by performing close content
analysis on the entire MiniPolymath 3 conversation and on sections of
MiniPolymath 1.  Again, this was conducted alongside discussion with
the other co-authors throughout the process.
A fourth co-author  (with a first degree in Mathematics)
later extended the tag set
with
additional informal logical relationships, such as \texttt{analogy}, and specific
content-focused relationships, such as \texttt{sums}, which played a
role in the further examples we treated in Section
\ref{iatc-examples}.  These extensions were again reviewed by all co-authors.

@@

The data that underlie this section were generated in a series of online experiments in collaborative problem solving convened by mathematician Terence Tao \citeyearpar{tao2009imo,tao2011imo}.
\thesis{We use IATC to expand on the analysis of
  this data presented by \citet{pease-and-martin},} showing how IATC can
advance the theory of mathematical argument through the
detailed analysis of real world examples, as per \citet{Carrascal2015}.

In the earlier analysis, Pease and Martin took a broad stroke approach, with each
comment comprising a single unit to be tagged.  They developed a typology
of five intuitive comment types, based on the mathematical content of each
comment: \emph{examples},
\emph{conjectures},
\emph{concepts},
\emph{proofs},
and \emph{other}. 
%%
In order to assign comments to these categories, both authors
performed close content analysis, together, on all comments posted
between the time at which Tao posted the problem to his blog
(8pm, UTC on July 19th, 2011) and the time he posted
that it had been solved (9.50pm, UTC on July 19th, 2011).
The discussion comprised 147 comments over 27 threads.
%%
Ten comments were assigned to more than one category.
(Both authors have a first degree in
Mathematics; one of us has a PhD in Mathematics and over 10 years
experience as a professional research mathematician; the other has a
PhD in a related discipline and more than 10 years experience studying
mathematical reasoning.)

Our present IATC analysis of the same data is designed to give a more complete picture of the
linguistic, dialectical, and inferential structure of the comments
that fall within the five intuitive categories mentioned.  There are three main
differences between the two analyses.  First, in comparison withh the earlier
broad-stroke analysis, the IATC analysis is richly detailed, with a
unit defined as any quantum of commentary with taggable
content.  Secondly, our focus in the earlier analysis was purely on
mathematical content, and on the {\em type} of mathematical
content in particular.  This contrasts with our
present analysis, in which we provide a more fine-grained representation of mathematical
content in the taggable units, and furthermore take into account
linguistic, dialectical, and inferential structure.
Third, the IATC analysis takes into consideration the entire
MiniPolymath 3 conversation, including the comments that came after
Tao had announced that a proof had been found.

The new analysis, accordingly, adds depth to our earlier analysis.
Crucially, the new perspective will be more relevant to argumentation
theorists, and supports a detailed understanding of what went on in
the process of constructing the collaborative proof.  The earlier
typology provided an initial way to sort the content, whereas the IATC
tag set developed along with our analysis via the iterative,
discursive method discribed in Section \ref{iatc}.  Though they cover
the same data and show some correlations, as described below,
the latter categorisation was not derived from the earlier one.

@@

\paragraph{Our main example in this section is MiniPolymath 3 (MPM3),
which we tagged into IATC in its entirety.}
(This work was carried out by one co-author with a first degree and PhD in Mathematics,
in consultation with others as described in Section \ref{iatc}.)
As an indicative sample, the first three comments and their tags are shown in Figure
\ref{fig:mpm3-example}.
Figure \ref{fig:grammar-categories} shows how tags from IATC's five grammatical
categories were distributed over time.  Thus, for example, we see
`\texttt{value}' tags used early in the discussion as strategies are
being considered, and again later in the discussion when solutions are
being vetted.
Figure \ref{fig:alternative} gives another view of the timeline,
showing how the comments were categorised into the 5-part typology
from Pease and Martin.  In the initial
categorisation developed for that paper, comments were allowed to be
in multiple categories at once.
\paragraph{Here, to facilitate a clean mapping to IATC, we redid the categorisation with the requirement that each comment should fit into exactly one main category.}
We arrived at a nearly equal division of comments among the five categories:
\emph{example} (20.3\%) \emph{conjecture} (21.2\%),
\emph{concept} (19.5\%), \emph{proof} (19.5\%), and \emph{other}
(19.5\%).  (This replication work was carried out independently by
one of the coauthors with a first degree in Mathematics.)

@@

\paragraph{One might suspect that \texttt{Suggest} should be used only within conjectures, but in the current categorisation it is used somewhat more frequently along with concepts.}
This is partly explained by the fact that \texttt{Suggest} can be used
to introduce either a \texttt{goal} or a \texttt{strategy}.  Sometimes goals
represent conceptual tidying, as in ``I guess there is an odd / even
number of point distinction to do'' \cite[\href{https://wp.me/pAG2F-41\#comment-3398}{July 19, 9:31 pm}]{tao2011imo}.
%%
\paragraph{Furthermore, despite of our self-imposed constraint to map each comment only to the most salient of the five categories, in practice a comment may simultaneously introduce a concept along with a conjecture that applies that concept.}
For example the straightforward concept of
``restriction[s] on how the next pivot is chosen''
appears along with the more speculative conjecture  ``Can we
start with a complete graph and all cycles on that graph and just
discard the ones that don't follow the restrictions to converge on the
ones that do?''
\cite[\href{https://wp.me/pAG2F-41\#comment-3368}{July 19, 8:56 pm}]{tao2011imo}.
The need to introduce concepts also applies in the case of more outlandish conjectures,
such as
``It might be fun to use projective duality''
\cite[\href{https://wp.me/pAG2F-41\#comment-3324}{July 19, 8:23 pm}]{tao2011imo}.
However, a concept may suggest a vague method
without raising a conjecture as such,
e.g., ``I'm thinking spirograph rather than convex hull''
\cite[\href{https://wp.me/pAG2F-41\#comment-3349}{July 19, 8:44 pm}]{tao2011imo}.

@@

\paragraph{There are other resources available which could further expand IATC's offerings in this regard.}
For example, a recent special issue of \emph{Argument \& Computation} \cite{ac-special-issue} includes papers detailing the usefulness of
rhetorical structures for argument mining.   \citet{mitrovic2017ontological}, in that volume, indicate the SALT Rhetorical Ontology \cite{groza2012advances} as relevant prior work.  SALT contains three categories---coherence relations, argument scheme relations, and rhetorical blocks---each of which unfolds with considerable further detail.  These three categories can be seen as somewhat analogous to IATC's grammatical categories.   \citet{mitrovic2017ontological} and \citet{lawrence2017harnessing} point to foundational work of \citet{fahnestock1999rhetorical,fahnestock2004figures} on the argumentative function of rhetorical figures, particularly in science writing.  IATC might be profitably connected to such analyses. Furthermore, the integration of rhetoric into argument mining highlights the relevance of structures that are rather different from the IAT-style transitions that have been used in work summarised by \citet{budzynska2015automatically}.  White's \citeyearpar[p.~6]{white1978tropics} pithy assertion that ``logic itself is merely a formalization of tropical strategies'' can serve as an additional provocation to develop structural analyses of this sort.
